Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-13 rebuttal.tex

\comment{Section 2.2. Structure Enumeration: it is not clear what this
  section is trying to convey to the reader \ldots}
\response{We have rewritten this section to simplify the language as
  well as be more explicit in applications of and challenges in this
  technique. We have also provide relevant references to key topics in
  this area}

\comment{p2, c1: I think the authors could benefit from a re-write of
  the section detailing molecular structure representations. \ldots
  Figure 1: what is meant by 'closer to reality' This needs to be
  clarified. This figure and its caption requires a little more
  consideration to appropriately convey the message, }
\response{We have updated the caption to remove the phrase ``closer to
  reality'' since this is not the main point. Instead we have noted
  that the representations contain differing amounts and types of
  information that are equally valid, but suited for different
  purposes. The text on page 2 has been updated to better represent
  this view}

\comment{p1, c2, l31: it is not clear to this reviewer what software
  and algorithms have only recently been made available. Certainly,
  the referenced book to which they refer contains a large number of
  algorithms that have already been published in the primary
  literature and frequently many decades ago. If the open source
  community has provided true novelty then please articulate this,
  otherwise cite the primary references that report these
  discoveries.}
\response{The text has been updated to note that most of the
  algorithms in cheminformatics have been known since the mid 20th
  century, but that freely accessible toolkits implementing them are
  relatively recent}

\comment{The paragraph in the second column discussing descriptors and
  the proper description requires rewriting too since it provides a
  very vague overview that is not particularly informative.}
\response{We have updated the discussion of descriptors to be more
  focused}

@@

\comment{P5, c1, l41: I would dispute the assertion that these methods
  "ignore reception interactions" since the aim is to identify a
  correlation of biological response (e.g. pIC50) with chemical
  structure. The biological response is an explicit measurement of
  receptor interactions on the protein.}
\response{We have updated
  the text to explicitly note that QSAR models do not usually take
  into account \emph{receptor features}. However we note that while
  receptor information is implicit in the $IC_{50}$, the value also
  includes other non-receptor related features such as
  permeability. Furthermore, lack of receptor information in QSAR
  models has been noted as the origin of activity cliffs (Guha and Van
  Drie, \textit{J.~Chem.~Inf.~Model.}, \textbf{2008}, \textit{48},
  1716--1728), thus supporting the statement that traditional QSAR
  models do indeed miss important information on receptor-ligand
  interactions.}

\comment{P5, c2, l22: I would mention naive Bayesian classifiers as well as
this is perhaps the most widely applied method in the field.}
\response{We have included the Na\"{i}ve Bayes as well}

\comment{P5, c2, l42: formal citation for one of Hopfinger?s papers is
  required here.}
\response{The relevant reference has been added}

\comment{P5, c2, l58: this assertion is made with no evidence to back
  it up. Why should multi-target models be more reliable? It is not
  clear to me that they should.}
\response{We have rephrased this statement to be less dogmatic. While
  it is true that for the case of linear models, a multivariate
  multiple regression is in general, equivalent to multiple individual
regression models, one could argue that taking into account the
correlation structure between the multiple y's could lead to }

### 2011-11-13 paper.tex
more updats to the text

The first step in the predictive modeling of biological activities is
to describe compounds in a way amenable to computerized data mining,
while at the same time capturing information relevant for the property
one attempts to model. Here multiple approaches exist: Either we
transform molecular graphs into a numerical vector representation
(a.k.a. \emph{descriptors} or \emph{features}) of chemical structures,
or we use algorithms for comparing molecular graphs directly
(\emph{molecule kernel}). In the first case thousands of descriptors
(many of which are equivalent) have been described
\cite{todeschini2000}, and a variety of tools are available for their
calculation. Given a set of molecules, observed activities and a
descriptor vector, the problem can be considered a traditional
statistical modeling problem. In the second case no explicit vector
representation is required, which prevents the challenge of
transforming the molecular graph into a vector without losing
information. On the other hand, the implicit kernel encoding imposes
the challenge that we have to compare all compounds against all
compounds, which imposes large-scale mining challenge, which risks
overrunning available processing time and space. Note that the public
PubChem dataset alone contains over 30 million chemical compounds and
over 70 million substances.

In order to judge which predictive modeling approach is applicable in
every case, we firstly have to consider the goal of a the model in a
cheminformatics setting. In some cases, pure predictive ability is
desired -- such as in a virtual screening setting. For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate the predicted
outputs into real-world compounds. In such cases, ``black-box
methods'' can be less useful, though attempts have been made to
interpret vectorial models built using neural networks, na\"{i}ve
Bayes, Support Vector Machines or random forests.

Secondly, the concept of model applicability -- the question of when
the prediction of the model for a new object is reliable -- has grown
in importance with the increasing use of predictive models in
regulatory settings. A variety of methods have been developed that
attempt to characterize the domain of a model. These methods not only
determine whether models are applicable, but can also be used for
deciding if additional biological experiments are required for
reducing the prediction error on certain compound classes.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Computer science techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

>>>>

The first step in the predictive modeling of biological activities is
to generate molecular descriptors (a.k.a, features) that are numerical
representations of different structural features. Here multiple
approaches exist. We can transform molecular graphs into a numerical
vector representation (a.k.a. \emph{descriptors} or \emph{features})
of chemical structures \cite{todeschini2000}, ranging from counts of
elements to eigenvalues of the Laplacian matrix.  Alternatively, we
can compare molecular graphs directly, via kernel methods. In these
cases, rather than compute vectorial representations, we directly
operate on the graph representations.

Both methods have advantages and disadvantages. The vectorial approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors. This is the feature selection problem and
is well discussed in the data mining literature. A kernel approach
does not require feature selection; but now, one faces the problem of
evaluating a data set in a pairwise fashion. Furthermore, one must now
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy Mercer's condition, and this is
not always possible with traditional cheminformatics based kernels
(such as a multiple common substructure based kernel). These
challenges can make kernel based methods prohibitive on very large
datasets.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we
looking for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
some of explanation as to why a molecule is predicted as toxic or
non-toxic, for example. The former scenario is common in virtual
screening settings, where we require high degree of accuracy and fast
predictions, but do not really care as to why one molecule is active
and another inactive. These types of models can be black boxes (neural
networks) and algorithmic models \cite{Breiman:2001fk} (random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we require to understand the effect of a
certain structural feature on the observed potency and expect that the
model will provide insight. These types of models are generally
distributional in nature (such as linear regression and PLS) though
some algorithmic approaches such as random forests can provide some
degree of insight.

Having chosen a modeling approach, a vital question is that of model
reliability which is closely tied to the concept of model
applicability -- the question of when the prediction of the model for
a new object is reliable. This issue has grown in importance with the
increasing use of predictive models in regulatory settings. A
misprediction (due to the model not bein applicable to a certain class
of inputs) can have significant financial repurcussions. A variety of
methods have been developed that attempt to characterize the domain of
a model. These methods not only determine whether models are
applicable, but can also be used for deciding if additional biological
experiments are required for reducing the prediction error on certain
compound classes.

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in
isolation. Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that such aspects be taking into account and, as
far as possible, take into account the receptor
simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.
